---
title: "Navigating Toronto's Homelessness Landscape: A Comprehensive Data Analysis of Trends, Compositional Shifts, and Projections for the City's Homeless Population Dynamics (2018-2023)"
author: 
  - Yiyi Feng, Yingxvan Sun
date: today
date-format: long
format:
  pdf:
    toc: true
    number-sections: true
number-sections: true
abstract: "In the context of economic struggles and the impact of the pandemic, this analysis deeply examines Toronto's shelter system from 2018 to 2023. It explores changes in the homeless population across population groups, genders, and age groups, studying population dynamics closely. A linear model was generated to predict the trend for the number of individuals using shelter systems in the future. The modeling predicts an increase in homelessness, emphasizing the need for proactive approaches. These discoveries reveal the necessity for more shelter facilities to address the growing challenges ahead."
bibliography: references11.bib
---


```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Writing a paper titled "Navigating Toronto's Homelessness Landscape: A Comprehensive Data Analysis of Trends, Compositional Shifts, and Projections for the City's Homeless Population Dynamics (2018-2023)"
# Author: Yiyi Feng, Linda Sun
# Date: 10 March 2024
# Contact: yiyi.feng@mail.utoronto.ca, lindayx.sun@mail.utoronto.ca

####Workspace setup ####
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggpubr)
library(kableExtra)
library(here)

#### Read data ####
cleaned_data <-
  read_csv(
    file = "../data/cleaned_data.csv",
    show_col_types = FALSE)

cleaned_data_figone <-
  read_csv(
    file = "../data/cleaned_data_fig1.csv",
    show_col_types = FALSE)

cleaned_data_figtwo <-
  read_csv(
    file = "../data/cleaned_data_fig2.csv",
    show_col_types = FALSE)

cleaned_data_figthree <-
  read_csv(
    file = "../data/cleaned_data_fig3.csv",
    show_col_types = FALSE)

cleaned_data_model <-
  read_csv(
    file = "../data/cleaned_data_model.csv",
    show_col_types = FALSE)

# Retrieve the linear model from the saved .rds file
lm_model <- readRDS(file = here("model", "linear_model_month.rds"))
```
# Introduction [^1]

Undoubtedly, homeless individuals in Toronto have been facing serious challenges in accessing nighttime accommodation in recent years. As early as 2020, the general manager of the city's shelter, support, and housing administration openly stated that an average of 72 people were being turned away from shelters due to overcrowding every day, a number that only continued to rise [@homelessinTor]. By the winter of 2022, this figure had surged to a staggering 168 people [@cbcnews]. For residents living in Toronto, this reality is apparent - despite shelter occupancy rates exceeding 98% [@streetneed], a significant number of individuals still resort to sleeping on the streets, a reality witnessed firsthand by residents and students alike.

Our understanding of the changing trend in the composition of the homeless community in Toronto is little. While the overall number of homeless individuals is documented, the specific demographic shifts within this population, such as variations in age groups and gender distribution, have not been comprehensively explored. Understanding these trends is essential for developing targeted interventions and resource allocation strategies to address the diverse needs of the homeless population in Toronto more effectively.

To do this, we analyzed Toronto's homeless population,  to understand the trends in this community, including changes in overall numbers and demographic composition. Leveraging data obtained from the Toronto Open Database, we aim to find the complexity and evolving trends of Toronto's homeless community. We analyze the status and demographic composition of Toronto's homeless population from 2018 to 2023, focusing on changes in status, such as the number of homeless individuals transitioning between housing and shelters, and long-term use of the shelter system, as well as demographic characteristics including gender, age, and identity.

A series of structured analyses were conducted. We begin by examining the annual counts of transitions of homeless individuals between housing, shelters, and other statuses within the Toronto shelter system from 2018 to 2023. Following this, we analyzed the composition of homeless population groups and the trend of age distribution within the homeless population. Similarly, the trend of gender composition across different population groups in the homeless community was also analyzed. Finally, we present the predictions from our linear model for future trends in the number of homeless individuals in Toronto.

Our analysis reveals several key findings. Firstly, the linear model we built suggests a steady increase in the number of homeless individuals in Toronto, indicating the need for expansion of the shelter system to accommodate more people. Additionally, our examination of the data through various figures provides insights into the nuanced dynamics of homelessness in the city. These include trends in the annual counts of homeless individuals transitioning within the Toronto shelter system, trends in the composition of homeless population groups and age distribution, and trends in gender composition across different population groups within the shelter system.

Understanding the trends and dynamics of homelessness in Toronto is of paramount importance, given its implications for public policy and social welfare. By finding out the challenges faced by the Toronto shelter system, we aim to inform policymakers and stakeholders in their efforts to address homelessness in the city.

[^1]: Please check https://github.com/kqlqkqlqF/Toronto-Shelter-System-Flow-between-2018-and-2023-with-Further-Predictions.git for more information.


# Data

## Data Source

The data for this analysis was collected from the shared Toronto Open Database [@opendatatoronto]. This study utilizes and analyzes the dataset titled "About Toronto Shelter System Flow". The dataset contains lots of information on the transition of homeless individuals in Toronto shelters from January 2018 to January 2024, including gender, age, quantity, and population groups individuals belong to (such as refugees). Additionally, the dataset provides information on the number of people leaving and entering shelter systems. It is released by the Toronto Shelter, Support, and Housing Administration, updated monthly, and has a high level of credibility. Moreover, the dataset has received full marks for freshness, metadata, accessibility, completeness, and usability on the Opendatatoronto website. Therefore, we consider the content of this dataset to be highly credible and utilize it as the primary data source for this paper. However, due to the limited data available for January 2024, we did not include this portion in the data analysis, as the smaller dataset size for 2024 may lead to unexpected analysis results.

## Features

Upon entering shelters, individuals utilizing shelter services are required to provide their name, age, gender, and group affiliation, which are recorded in the database. This dataset only records homeless individuals using overnight shelter services and does not include those utilizing other welfare policies, such as receiving free food or vaccinations. In the data, homeless individuals are divided into five age groups: under 16, 16 to 24, 25 to 44, 45 to 64, and over 65. Gender is categorized as male, female, and transgender/non-binary/two-spirit. The population group includes chronic, refugees, families, youth, single adults, non-refugee, and indigenous. however, the indigenous group was only included in the statistics starting from January 2022, as the authors of the dataset stated their intention to collect more detailed data, hence adding the subdivision. Here, chronic refers to homeless individuals who have continuously used shelter services for more than 180 days. Regarding the documentation of homeless transitions, the dataset provides six subdivisions: newly identified, return from housing, return to shelter, moved to housing, became inactive, and actively homeless. Newly identified refers to people who entered the shelter system for the first time; returned from permanent housing refers to people who previously used the shelter system, then moved to permanent housing, and have now returned; returned to shelter refers to people who were previously using the shelter system, then did not use the system for 3 months or longer, and have now returned; moved to permanent housing refers to people who were using the shelter system and have moved to permanent housing; became Inactive refers to people who last accessed shelter services three months ago; actively homeless refers to people who have used shelter services at least one time in the past three months and have not moved to permanent housing.

## Methodology

The data analysis was conducted using R [@citeR], a versatile statistical programming language. We utilized a range of packages to enhance our analysis. The tidyverse [@tidy] suite of packages provided a comprehensive toolkit for efficient data manipulation and visualization. Package ggplot2 [@gg2] allowed us to create compelling visualization. The here [@here] package simplified file management within our project directory structure. Additionally, kableExtra [@kable] was employed to generate visually appealing and customizable tables, enhancing the presentation of our findings. For Bayesian analysis, we utilized the rstanarm [@rstanarm] package, which provided an elegant interface to Stan, a cutting-edge platform for statistical modeling and computation. This allowed us to estimate relationships within our data using a Bayesian framework, providing valuable insights into our research questions. Report generation was seamlessly managed using knitr [@knitr], enabling the integration of R code within our document.  Other essential packages included tibble [@tibble], stringr [@stringr], lubridate [@lubridate], janitor [@janitor], and testthat [@testthat], each contributing to various aspects of our data analysis process, from data manipulation to quality assurance.

Due to the clarity of the data itself, our data cleaning process primarily focused on converting the raw data dates into the yyyy-mm-dd format and selecting the data needed for producing each figure respectively to ensure code was organized and minimize the amount of code in the final QMD file. For the first chart, intended to reveal Trends in the Annual Counts of Homeless Individuals Transitioning within the Toronto Shelter System between 2018 and 2023, we retained the data for the six transition status categories from monthly data and aggregated data within the same year for ease of subsequent chart generation. For the second chart, revealing Trends in the Composition of Homeless Population Groups and Age Distribution within the Toronto Shelter System between 2018 and 2023, we removed all data except for age and population groups. Similarly, for the third chart, Trends in Gender Composition of Homeless Population Across Different Population Groups within the Toronto Shelter System between 2018 and 2023, we retained only gender and group affiliation data. The final linear model utilized the cleaned dataset for the second figure, and additional cleaning was performed. It is important to note that we excluded data from January 2024 in all chart data to avoid errors.

# Model

## Model Setup

Our objective is to investigate how the number of individuals using Toronto's overnight shelter system has changed over time since January 2018 and to generate a relevant model to predict trends in this relationship. To achieve this goal, we first quantify time as the number of months elapsed since January 2018. For instance, data from February 2019 would be considered the data for the 13th month. Subsequently, we conducted a linear regression model for the total number of individuals using Toronto's overnight shelter system for each month since January 2018. In this model, the total number of individuals using shelters in a given month is represented by $y_i$, while the number of months elapsed since January 2018 is represented by $x_i$ for each month $i$:

\begin{align}
y_i &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{AGE}_i \\
\beta_0 &\sim \mbox{Normal}(90, 10) \\
\beta_1 &\sim \mbox{Normal}(-0.1, 0.05)
\sigma &\sim \mbox{Exponential}(1)
\end{align}

The analysis was conducted in R using the rstanarm package, which facilitates the integration of prior beliefs about parameters and the estimation of posterior distributions. We applied normal prior distributions centered around our initial estimates to the intercept ($\beta_0$) and slope ($\beta_1$) parameters, informed by exploratory data analysis. Our choice of priors aimed to capture reasonable assumptions about parameters before observing the data.

The prior for the intercept reflects an expectation that newly constructed buildings typically receive high safety scores around 90, with a broad range to accommodate uncertainty. Similarly, the prior slope suggests a minor negative impact of several months since 2018 January on the total number of individuals who used the shelter system in Toronto, acknowledging some uncertainty in the precise effect. Additionally, we assigned an exponential prior with a rate of 1 to the standard deviation ($\sigma$), reflecting a belief in the likelihood of smaller variability in evaluation scores. The Bayesian regression was performed using the stan_glm function from rstanarm, enabling inference that leverages both the dataset and our prior knowledge. This function effectively combined the priors with the dataset to derive posterior distributions for the model parameters.

## Model Justification

We assume that the coefficient ($\beta_1$) of the number of months passed since January 2018 will be positive, reflecting a trend towards increasing overcrowding in Toronto shelters. This expectation stems from various changes observed between 2018 and 2023: firstly, the disproportionate rise in housing prices compared to wages has made it increasingly difficult for people to secure permanent accommodation; secondly, the economic impact of the COVID-19 pandemic has severely affected the financial status of many individuals. These changes are likely to increase the homeless population, leading to a rise in the number of people utilizing shelter overnight services. Therefore, the slope ($\beta_1$) is our primary focus for validation, as it will determine whether we anticipate more or fewer people to use shelter overnight services in 2024. Additionally, we believe the intercept ($\beta_0$) will also be positive because since January 2018, many people have been using the shelter system. Thus, regardless of whether the outcome is positive or negative growth, the intercept will be a positive number.

\newpage
# Result

## Trends in Homelessness Transitioning within the Toronto Shelter System
```{r}
#| label: tbl-one
#| tbl-cap: Annual Counts of Homeless Individuals Transitioning between Housing, Shelters, and Other Status Within the Toronto Shelter System between 2018 and 2023
#| echo: false

# Create a table plot using kableExtra with adjusted width
# Generate new column names with line breaks
colnames(cleaned_data_figone) <- c("Year", 
                                   "Returned from Housing", 
                                   "Returned to Shelter", 
                                   "Newly Identified", 
                                   "Moved to Housing", 
                                   "Became Inactive", 
                                   "Actively Homeless")

# Create a table plot using kableExtra
kable(cleaned_data_figone, align = "c", col.names = colnames(cleaned_data_figone), linesep = "\n")
```

```{r}
#| label: fig-one
#| fig-cap: Trends in the Annual Counts of Homeless Individuals Transitioning within the Toronto Shelter System between 2018 and 2023 
#| echo: false
#| message: false
#| warning: false

#### Generate the first figure ####
# Convert year to factor to maintain ordering on x-axis
cleaned_data_figone$year <- factor(cleaned_data_figone$Year)

# Reshape data from wide to long format for ggplot
cleaned_data_figone_long <- tidyr::pivot_longer(cleaned_data_figone, 
                                                cols = -year, 
                                                names_to = "Category", 
                                                values_to = "Value")

# Remove rows where "year" appears in the "Category" column
cleaned_data_figone_long <-  cleaned_data_figone_long[cleaned_data_figone_long$Category != "Year", ]

# Define the themes
theme_minimal_custom <- theme_minimal() +
  theme(axis.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Plot without year data on x-axis
ggplot(cleaned_data_figone_long, aes(y = Value, color = Category, group = Category)) +
  geom_smooth(aes(x = year), method = "loess", se = FALSE, size = 1, alpha = 0.8) +
  geom_point(aes(x = year), size = 3, alpha = 0.8) +
  scale_color_viridis_d() +
  labs(x = "Year",
       y = "Population Count",
       color = "Category") +
  theme_minimal_custom +
  theme(panel.background = element_rect(fill = "#f2eada"),  #set background color
        axis.text = element_text(face = "bold"),              #bold axis text
        axis.title = element_text(face = "bold"),
        panel.grid.major = element_line(color = "gray"),      #set color of major grid lines
        panel.grid.minor = element_line(color = "lightgray")) #set color of minor grid lines
```

The data provided in [@tbl-one] and [@fig-one] offers insight into the dynamic changes in the homeless population within the Toronto shelter system from 2018 to 2023. 

A notable trend observed is the fluctuation in the annual count of newly identified individuals entering the shelter system. Peaking in 2018 and gradually declining over the following years, the trend suggests a persistent need for shelter services, with a gradual increase observed from 2020 onwards.
Another significant aspect of the data is the return of individuals from permanent housing to the shelter system. This phenomenon is concerning as it indicates potential housing instability for the population. Fortunately, the numbers have remained relatively stable from 2018 to 2023, and they are lower than those transitioning to permanent housing each year. However, the declining trend in individuals moving to stable housing from 2018 to 2021 is worrisome. Although there is a slight increase in 2022 and 2023, it remains below the 2018 levels, suggesting a decrease in individuals transitioning to permanent housing.
The data also captures changes in individuals returning to shelters and those becoming inactive within the shelter system. While there is a gradual decline in the number of individuals becoming inactive homeless over the five years, the decrease is minimal, similar to the trend in individuals returning to shelters. Without further explanation, it's challenging to draw reliable conclusions on the decrease in the number of inactive homeless individuals or the reasons behind individuals returning to shelters since their experience after leaving the shelter system is unknown. They could be moving to other cities, finding housing, or dying. Therefore, if there's no further detail for the reason they left the shelter system, it is hard to understand the implications of the data fully, and no conclusion can be drawn.

Lastly, individuals from the actively homeless category highlight the ongoing demand for shelter services and support. Although there was a decline in the number of actively homeless individuals in 2020 and 2021, it surged to new highs in 2022 and 2023. This suggests a significant increase in the number of individuals utilizing overnight shelter services in the past two years, indirectly indicating a potential increase in Toronto's homeless population from 2022 to 2023.
In conclusion, these data paint a complex picture of homelessness within the Toronto shelter system, with a trend of increasing actively homeless individuals. 

## Composition of Homeless Population Groups and Correcponding Age Distribution
```{r}
#| label: fig-two
#| fig-cap: Composition of Homeless Population Groups and Age Distribution within the Toronto Shelter System between 2018 and 2023
#| echo: false
#| message: false

#### generate the 2nd graph ####
#filter for necessary population groups
cleaned_data_figtwo <- cleaned_data_figtwo %>%
  filter(population_group %in% c("All Population", "Chronic", "Refugees", "Families", "Indigenous", "Non-refugees"))

#reshape the data for graph generating
reshaped_data <- cleaned_data_figtwo %>%
  pivot_longer(cols = ageunder16:age65over, names_to = "age_group", values_to = "count")

#define a custom color palette for better look of the graph
starry_night_palette <- c("#145b7d", "#70a19f", "#008792","#00ae9d", "#76becc", "#94d6da")

#plot the data
ggplot(reshaped_data, aes(x = age_group, y = count, fill = population_group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.75) +
  labs(x = "Age Group", y = "Population Count", fill = "Population Group") +
  scale_fill_manual(values = starry_night_palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Before beginning the analysis of [@fig-two], several points need to be explained. Firstly, within the population groups depicted in this graph, apart from refugees and non-refugees, the other options are not mutually exclusive. This means that a person can simultaneously belong to the chronic, families, and refugee groups, but cannot be a refugee and a non-refugee at the same time. This explains why the sum of individuals in these population groups exceeds the number in all population categories. Secondly, as mentioned earlier, the number of individuals in the indigenous group was only recorded starting in 2020, so the actual number of indigenous individuals who have used the shelter system should be higher than what is shown in the graph.

Individuals aged between 25 and 44 constitute the largest portion of the homeless population, while those aged between 45 and 64 represent the second largest group, and those aged 65 and above represent the smallest group. From the perspective of population groups, non-refugees constitute the largest group, followed by refugees and then chronic individuals, with indigenous individuals being the smallest group. From this distribution trend, we can infer the following information:

Firstly, except for the age group under 16, chronic individuals constitute a significant portion of the homeless population in other age groups, indicating that a considerable number of homeless individuals rely on overnight shelter services for an extended period and find it difficult to secure permanent housing, thus perpetuating their homelessness.

Secondly, although refugees constitute a smaller proportion of the homeless population compared to non-refugees overall, in the age group under 16, the number of refugees far exceeds the proportion of non-refugees. This suggests that there are a significant number of parenting teenagers or even younger children among the refugee population who are homeless. Additionally, we observe that the proportion of families in the age group under 16 also significantly increases, slightly surpassing the proportion of refugees and almost equaling the "all population" category. This evidence confirms our previous conclusions and supports the notion that these young refugee homeless individuals are likely to be wandering alongside their refugee family members.

Thirdly, unlike refugees, homeless individuals categorized as non-refugees are mainly distributed in the age group of 25 and above. Particularly in the age group of 65 and above, the number of non-refugees is almost equivalent to the "all population" category.

In summary, chronic individuals are prominent across age groups, highlighting challenges in securing permanent housing. Refugees tend to enter the shelter system as a whole family with their young children, indicating a concerning trend of homelessness among refugee families with children, while the non-refugees dominate the older age groups. 

## Gender Composition Across Homeless Population Groups

```{r}
#| label: fig-thr
#| fig-cap: Trends in Gender Composition of Homeless Population Across Different Population Groups within the Toronto Shelter System between 2018 and 2023
#| echo: false
#| message: false

#### generate the 3rd graph ####

# Rename the columns first
cleaned_data_figthree_renamed <- cleaned_data_figthree %>%
  rename(
    "Year" = year,
    "Population Group" = population_group,
    "Male" = gender_male,
    "Female" = gender_female,
    "Transgender/Non Binary/Two spirit" = gender_transgender_non_binary_or_two_spirit
  )

# Reshape the data for graphing
cleaned_data_figthree_long <- tidyr::pivot_longer(cleaned_data_figthree_renamed, 
                                                  cols = c(Male, Female, `Transgender/Non Binary/Two spirit`),
                                                  names_to = "gender",
                                                  values_to = "count")

# Plot the data
ggplot(cleaned_data_figthree_long, aes(x = Year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_brewer(palette = "Paired") + 
  labs(x = "Tear",
       y = "Population Count",
       fill = "Gender") +
  scale_y_continuous(labels = scales::comma) +  #format y-axis labels with commas
  theme_minimal() +
  facet_wrap(~ `Population Group`, ncol = 3, scales = "free_y") +  #facet by population groups with 3 columns
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7),  # set y-axis label size and make it bold#rotate x-axis labels
        strip.text = element_text(size = 8),  #adjust the size of facet labels
        legend.position = "bottom",  #position legend at the bottom
        legend.box.spacing = unit(0.3, "lines"),  #set spacing around legend
        legend.spacing = unit(0.3, "lines"))  #set spacing within legend
```

The data presented in [@fig-thr] provides insights into the gender dynamics within Toronto's homeless population across different population groups from 2018 to 2023. A consistent trend observed throughout the years is the higher representation of males compared to females and individuals identifying as transgender, non-binary, or two-spirit. This imbalance suggests a prevalent gender disparity among those experiencing homelessness, with males comprising a larger proportion of the homeless population.

However, within specific population groups, variations in gender composition emerge. For instance, in the chronic population group, which likely encompasses individuals experiencing long-term homelessness, both males and females are prominently represented, indicating a diverse demographic within this subgroup. In contrast, the families population group, which likely consists of homeless families with children, demonstrates a higher count of females compared to males and individuals identifying as transgender, non-binary, or two-spirit. This gender distribution within homeless families may be influenced by factors such as caregiving responsibilities and access to support services, indicating the unique dynamics within this subgroup. The "Refugee" population group exhibits a relatively balanced distribution between males and females, with a smaller count of individuals identifying as transgender, non-binary, or two-spirit. This suggests a diverse composition within refugee populations experiencing homelessness, possibly influenced by factors such as migration patterns and resettlement challenges.

Furthermore, the data reveal temporal fluctuations in the chronic population group, which demonstrates a consistent increase in count from 2018 to 2023. This upward trend suggests a growing number of individuals experiencing long-term homelessness.

## Result of the Model

### Model Coefficients

```{r}
#| label: coe
#| tbl-cap: Summary Statistics for the Coefficients of the Linear Model
#| echo: false

# Extract coefficients from the Bayesian model
coefficients <- coef(lm_model)

# Convert coefficients to a data frame for better formatting
coefficients_df <- data.frame(
  Term = c("Intercept", "Slope"),
  Estimate = unname(coefficients) 
)

# Round the estimates to three digits
coefficients_df$Estimate <- round(coefficients_df$Estimate, 3)

# Create a nicely formatted table
coefficient_table <- kable(coefficients_df,
                           col.names = c("Term", "Estimate"),
                           align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center")

# Print the summary table
coefficient_table
```

**Intercept** ($\beta_0$)  : The intercept represents the estimated number of the total population that used shelter overnight service in Toronto at the beginning of the observation period, which is 2018 January, which is 8758.35

**Slope** ($\beta_1$)  : The slope quantifies the average change in the number of the total population who used shelter overnight service in Toronto per month. It is expected that 14.605 more individuals are using this service per month.

**Explanation**:
The analysis showed a positive correlation between the number of people using shelter overnight services each month and the number of months that have elapsed since January 2018. Estimates indicate that as time passes from January 2018, more individuals use shelter overnight services.

### Model Equation

\begin{align}
y_i &= 8758.35 + 14.605 x_i + \varepsilon_i
\end{align}

### Residuals

```{r}
#| echo: false

# Extract residuals from the model
residuals <- resid(lm_model)
summary(residuals)
```

**1. Minimum Residual**: The minimum residual is -1826.5, indicating that there is at least one observation for which the model's predicted total population was lower than the actual population by approximately 1826.5 individuals. This suggests the presence of potential outliers or factors not accounted for by the model.

**2. First Quartile Residual**: The first quartile (Q1) of the residuals is -655.6, implying that 25% of the observations have total populations that are less than 655.6 individuals below the model's prediction.

**3. Median Residual**: The median residual is 226.4, indicating that half of the observations have residuals below 226.4, suggesting a slight positive bias in the model's predictions. This means that, on average, the model tends to slightly underestimate the total population.

**4. Third Quartile Residual**: The third quartile (Q3) is 712.4, signifying that 75% of the observations have total populations within 712.4 individuals of the model's predictions or better. This suggests that the model provides reasonably accurate predictions for the majority of observations.

**5. Maximum Residual**: The maximum residual is 1185.1, suggesting that there is at least one observation for which the model's predicted total population was higher than the actual population by approximately 1185.1 individuals. Similar to the minimum residual, this could indicate the presence of outliers or unaccounted factors.

**Interpretation**: The distribution of residuals provides insights into the model's performance and the presence of potential outliers or discrepancies between the model's predictions and the actual observations. Further investigation into the reasons behind significant deviations can help refine the model and improve the accuracy of its predictions. Additionally, assessing the model's assumptions, such as linearity, homoscedasticity, and normality of residuals, is essential to ensure the validity of the statistical inferences drawn from the model.

### Graphing the Model
```{r}
#| label: fig-fou
#| fig-cap: The Linear Relationship between Number of Month Passed and the Number of individuals use Shelter Overnight Service
#| echo: false
#| message: false

# Create a scatter plot with a regression line to visualize the relationship
ggplot(cleaned_data_model, aes(x = month_number, y = total_population)) +
  geom_point(alpha = 0.7, size = 3, color = "darkblue") +  # Adjust transparency, size, and color of points
  geom_smooth(method = "lm", color = "darkorange", size = 1.5, linetype = "dashed") +  # Change the size and type of the regression line
  labs(
    x = "Number of Months Passed Since 2018 January",
    y = "Total Homeless Population",
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10), # Rotate x axis text and adjust horizontal justification
    axis.text.y = element_text(size = 10),
    legend.position = "none", # Hide the legend if not necessary
    plot.margin = margin(1, 1, 1, 1, "cm"), # Adjust the margin around the plot
    panel.grid.major = element_line(color = "lightgray", linetype = "dotted"), # Add dotted grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.line = element_line(color = "black"), # Change axis line color
    axis.title = element_text(size = 12), # Adjust axis title size
    plot.title = element_text(size = 14, face = "bold") # Adjust plot title size and style
  ) +
  coord_cartesian(ylim = c(5000, 12000)) # Set the y-axis limits

```

@fig-fou visualizes the relationship between the number of months passed since January 2018 and the total number of individuals utilizing the overnight shelter service in Toronto. The blue dashed line represents the linear regression line fitted to the data, capturing the average trend in the homeless population over time.

# Discussion

## Summery of Findings

Our key findings underscore an alarming rise in homelessness, notable shifts in demographic composition, and a gender imbalance, all highlighting the multifaceted nature of homelessness.

The analysis of homeless population dynamics in Toronto from 2018 to 2023 revealed fluctuating trends in shelter entry, with a peak in 2018 followed by a gradual decline and a slight increase from 2020 onwards. The data also highlighted potential housing instability through individuals returning from permanent housing to shelters, although overall stability was observed.  Chronic individuals were prevalent across age groups, indicating persistent challenges in securing permanent housing. Refugee families, often with young children, contributed to homelessness trends, while non-refugees dominated older age groups. Gender disparities were evident, with males comprising a larger proportion of the homeless population, especially among chronic individuals, while females were more represented in homeless families.

The estimation results provide insights into the magnitude of homelessness, with the intercept ($\beta_0$) indicating an estimated total homeless population of 8758.35 individuals in January 2018. The positive slope ($\beta_1$) signifies an average monthly increase of approximately 14.605 individuals since January 2018. These coefficients exhibit statistical significance, supported by small standard errors and highly significant p-values ($p < 0.001$), providing robust evidence for our estimates. However, the modest adjusted $R^2$ value suggests that only a small proportion of the variability in the homeless population can be explained by the model. Further analysis and consideration of additional factors are warranted to enhance the model's predictive accuracy and deepen our understanding of homelessness dynamics in Toronto.

Through predictive modeling, we have illuminated a critical need for expanding shelter infrastructure and implementing targeted support services. 

## Limitations

Although the analysis provides insights into the dynamics of homelessness in Toronto, there are still some limitations to consider. Firstly, these data primarily reflect individuals utilizing shelter services and may not capture the entire homeless population, including those living on the streets or in unstable housing conditions. Additionally, our data have limitations. For example, we cannot track the whereabouts of all homeless individuals who leave the shelter system; the data only record numbers without monitoring their activities. The reasons for individuals becoming inactive homeless may vary significantly, such as death, disappearance, or relocation, leading to confusion and ambiguity in predicting future dynamics. Thirdly, our dataset is based on the trends of homelessness from 2018 to 2023, which were affected by the multifaceted impacts of the COVID-19 pandemic, including but not limited to: increased difficulty in data collection due to social distancing measures, the health, economic, and social impacts of COVID-19 on the homeless population, and the potential closure of shelters due to the spread of the disease. These factors may reduce the effectiveness and accuracy of data collection. Moreover, we can observe a significant decrease in shelter utilization in 2020 and 2021 during the pandemic, which contradicts our initial assumptions. Therefore, we have some reservations about the reliability of the data during this period.

## Future Study

The findings from our study on homelessness in Toronto provide implications for both policy and future research. As the homeless population continues to grow steadily and faces challenges in securing permanent housing, urgent government intervention is needed to expand housing capacity, reduce housing costs, and enhance support services. Future research should explore the root causes of homelessness transitions, identify effective interventions to prevent homelessness recurrence, and evaluate the long-term impacts of housing policies and support programs. By understanding the underlying factors driving homelessness and implementing targeted interventions, it is possible to protect people from being homeless. As we move forward, the impact of the COVID-19 pandemic on homelessness dynamics is expected to diminish. Future research will provide access to post-pandemic data, reducing its variable impact on long-term homelessness trends. Furthermore, incorporating a wider range of comprehensive factors into our studies, including race and immigration status, will provide a deeper understanding of the social impacts of these factors on homelessness. Additionally, there is a need for the government to establish a more comprehensive measurement system for data collection, including tracking the dynamics of inactive homeless individuals and conducting categorized surveys of homeless families. This will enable a more precise assessment of the challenges faced by the homeless population and facilitate the development of targeted and effective measures to address them.


\newpage
# Bibliography








