---
title: "Navigating Toronto's Homelessness Landscape: A Comprehensive Data Analysis of Trends, Compositional Shifts, and Projections for the City's Homeless Population Dynamics (2018-2023)"
author: 
  - Yiyi Feng, Yingxvan Sun
date: today
date-format: long
format:
  pdf:
    toc: true
    number-sections: true
number-sections: true
abstract: "Against the backdrop of economic challenges and the ramifications of the pandemic, this analysis offers an intricate exploration of Toronto's shelter system from 2018 to 2023. Unveiling nuanced shifts in the homeless population's composition across single adults, families, genders, and age groups, the study scrutinizes population dynamics. It elucidates fluctuations in the actively homeless count and discerns patterns of return to homelessness. Notably, predictive modeling forecasts a concerning surge in homelessness, underscoring the urgency for proactive strategies. These findings illuminating the pressing need for expanded shelter infrastructure to meet the escalating challenges ahead."
bibliography: references11.bib
---


```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Writing a paper titled "Navigating Toronto's Homelessness Landscape: A Comprehensive Data Analysis of Trends, Compositional Shifts, and Projections for the City's Homeless Population Dynamics (2018-2023)"
# Author: Yiyi Feng, Linda Sun
# Date: 10 March 2024
# Contact: yiyi.feng@mail.utoronto.ca, lindayx.sun@mail.utoronto.ca

####Workspace setup ####
library(tidyverse)
library(janitor)
library(knitr)
library(ggplot2)
library(ggpubr)
library(kableExtra)

#### Read data ####
cleaned_data <-
  read_csv(
    file = "../data/cleaned_data.csv",
    show_col_types = FALSE)

cleaned_data_figone <-
  read_csv(
    file = "../data/cleaned_data_fig1.csv",
    show_col_types = FALSE)

cleaned_data_figtwo <-
  read_csv(
    file = "../data/cleaned_data_fig2.csv",
    show_col_types = FALSE)

cleaned_data_figthree <-
  read_csv(
    file = "../data/cleaned_data_fig3.csv",
    show_col_types = FALSE)
```
# Introduction [^1]

Undoubtedly, homeless individuals in Toronto have been facing serious challenges in accessing nighttime accommodation in recent years. As early as 2020, the general manager of the city's shelter, support, and housing administration openly stated that an average of 72 people were being turned away from shelters due to overcrowding every day, a number that only continued to rise. （1）By the winter of 2022, this figure had surged to a staggering 168 people.（2） For residents living in Toronto, this reality is starkly apparent - despite shelter occupancy rates exceeding 98%, a significant number of individuals still resort to sleeping on the streets or in parks, a reality witnessed firsthand by residents and students alike.

Despite significant efforts to address homelessness, there remains a notable gap in our understanding of the changing trend in the composition of the homeless community in Toronto. While the overall number of homeless individuals is documented, the specific demographic shifts within this population, such as variations in age groups and gender distribution, have not been comprehensively explored. Understanding these trends is essential for developing targeted interventions and resource allocation strategies to effectively address the diverse needs of the homeless population in Toronto.

To do this, we seek to conduct an in-depth analysis and study of Toronto's homeless population, aiming to understand the recent trends in this community, including changes in overall numbers and demographic composition. Leveraging data obtained from the Toronto Open Database, we aim to delve into the complexity and evolving trends of Toronto's homeless community. We analyze the status and demographic composition of Toronto's homeless population from 2018 to 2023, focusing on changes in status, such as the number of homeless individuals transitioning between housing and shelters, and long-term use of the shelter system, as well as demographic characteristics including gender, age, and identity.

To address this gap, we present a series of structured analyses of our findings. We begin by examining the annual counts of transitions of homeless individuals between housing, shelters, and other status within the Toronto shelter system from 2018 to 2023. Following this, we delve into the composition of homeless population groups and the trend of age distribution within the homeless population. Similarly, the trend of gender composition across different population groups in the homeless community was also analyzed. Finally, we present the predictions from our linear model for future trends in the number of homeless individuals in Toronto.

Our analysis reveals several key findings. Firstly, the linear model we built suggests a steady increase in the number of homeless individuals in Toronto, indicating the need for expansion of the shelter system to accommodate more people. Additionally, our examination of the data through various figures provides insights into the nuanced dynamics of homelessness in the city. These figures include trends in the annual counts of homeless individuals transitioning within the Toronto shelter system, trends in the composition of homeless population groups and age distribution, and trends in gender composition across different population groups within the shelter system.

Understanding the trends and dynamics of homelessness in Toronto is of paramount importance, given its implications for public policy and social welfare. By shedding light on the challenges faced by the homeless population and providing evidence-based insights, our study aims to inform and guide policymakers and stakeholders in their efforts to address homelessness in the city.

[^1]: Please check https://github.com/kqlqkqlqF/Toronto-Shelter-System-Flow-between-2018-and-2023-with-Further-Predictions.git for more information.


# Data

## Data Source

The data for this analysis was collected from the shared Toronto Open Database [@opendatatoronto]. This study utilizes and analyzes the dataset titled "About Toronto Shelter System Flow". The dataset contains a wealth of information on the movement of homeless individuals in Toronto shelters from January 2018 to January 2024, including gender, age, quantity, and group affiliation (such as refugees). Additionally, the dataset provides information on the number of people leaving and entering shelters. It is released by the Toronto Shelter, Support, and Housing Administration, updated monthly, and has a high level of credibility. Moreover, the dataset has received full marks for freshness, metadata, accessibility, completeness, and usability on the Opendatatoronto website. Therefore, we consider the content of this dataset to be highly credible and utilize it as the primary data source for this paper. However, due to the limited data available for January 2024, we did not include this portion in the data analysis, as the smaller dataset size for 2024 may lead to unexpected analysis results.

## Features

Upon entering shelters, individuals utilizing shelter services are required to provide their name, age, gender, and group affiliation, which are recorded in the database. It is important to note that this dataset only records homeless individuals using overnight shelter services and does not include those utilizing other welfare policies, such as receiving free food or vaccinations.In the data, homeless individuals are divided into five age groups: under 16, 16 to 24, 25 to 44, 45 to 64, and over 65. Gender is categorized as male, female, and non-binary. Group affiliation includes chronic, refugees, families, youth, single adult, non-refugee, and indigenous. It is noteworthy that the indigenous group was only included in the statistics starting from January 2022, as the authors of the dataset stated their intention to collect more detailed data, hence adding the subdivision. Here, Chronic refers to homeless individuals who have continuously used shelter services for more than 180 days. Additionally, by dividing the specific group's number of a given month by the total number of people using shelter services overnight, the dataset also records the percentage of each group in the overall population. Regarding the documentation of homeless transitions, the dataset provides six data points: new identified, return from housing, return to shelter, moved to housing, became inactive, and actively homeless. Newly Identified refers to people who entered the shelter system for the first time; Returned from Permanent Housing refers to people who previously used the shelter system, then moved to permanent housing, and have now returned; Returned to Shelter refers to people who were previously using the shelter system, then did not use the system for 3 months or longer, and have now returned; Moved to Permanent Housing refers to people who were using the shelter system and have moved to permanent housing; Became Inactive refers to people who last accessed shelter services three months ago; Actively Homeless refers to people who have used shelter services at least one time in the past three months and have not moved to permanent housing. 

## Methodology

The cleaned data were analyzed and performed using R [@citeR] with `tidyverse` [@citeTidyverse], `here` [@citeHere], `rstanarm` [@citeRstanarm], `modelsummary` [@citeModleSummary], `ggplot2` [@citeGgplot2], `knitr` [@citeKnitr], `marginaleffects` [@citeMarginalEffects], `plotly` [@citePlotly], `tibble` [@citeTibble], `margins` [@citeMargins], `testthat` [@citetestthat] and `kableExtra` [@citeKableExtra].

Due to the clarity of the data itself, our data cleaning process primarily focused on converting the raw data dates into the yyyy-mm-dd format, without making extensive changes to the overall dataset. Following the completion of general data cleaning, we performed separate data cleaning for each chart to ensure code were organized and minimize the amount of code in the final QMD file. For the first chart, intended to reveal Trends in the Annual Counts of Homeless Individuals Transitioning within the Toronto Shelter System between 2018 and 2023, we retained only a portion of the data for the six transition status categories from monthly data and aggregated data within the same year for ease of subsequent chart generation. For the second chart, aimed at revealing Trends in Composition of Homeless Population Groups and Age Distribution within the Toronto Shelter System between 2018 and 2023, we removed all data except for age and group affiliation. Similarly, for the third chart, which aimed to reveal Trends in Gender Composition of Homeless Population Across Different Population Groups within the Toronto Shelter System between 2018 and 2023, we retained only gender and group affiliation data. The final linear model utilized the overall dataset after data cleaning, and no additional cleaning was performed. It is important to note that we excluded data from January 2024 in all chart data to avoid errors.

# Result

## Trends in Homelessness Transitioning within the Toronto Shelter System
```{r}
#| label: fig-tab-one
#| fig-cap: Number of Homeless Individuals Transitioning between Housing, Shelters, and Other Status Within the Toronto Shelter System between 2018 and 2023
#| echo: false

# Create a table plot using kableExtra with adjusted width
# Generate new column names with line breaks
colnames(cleaned_data_figone) <- c("Year", 
                                   "Returned from Housing", 
                                   "Returned to Shelter", 
                                   "Newly Identified", 
                                   "Moved to Housing", 
                                   "Became Inactive", 
                                   "Actively Homeless")

# Create a table plot using kableExtra
kable(cleaned_data_figone, align = "c", col.names = colnames(cleaned_data_figone), linesep = "\n")
```

```{r}
#| label: fig-one
#| fig-cap: Trends in the Annual Counts of Homeless Individuals Transitioning within the Toronto Shelter System between 2018 and 2023 
#| echo: false
#| message: false
#| warning: false

#### Generate the first figure ####
# Convert year to factor to maintain ordering on x-axis
cleaned_data_figone$year <- factor(cleaned_data_figone$Year)

# Reshape data from wide to long format for ggplot
cleaned_data_figone_long <- tidyr::pivot_longer(cleaned_data_figone, 
                                                cols = -year, 
                                                names_to = "Category", 
                                                values_to = "Value")

# Remove rows where "year" appears in the "Category" column
cleaned_data_figone_long <-  cleaned_data_figone_long[cleaned_data_figone_long$Category != "Year", ]

# Define the themes
theme_minimal_custom <- theme_minimal() +
  theme(axis.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Plot without year data on x-axis
ggplot(cleaned_data_figone_long, aes(y = Value, color = Category, group = Category)) +
  geom_smooth(aes(x = year), method = "loess", se = FALSE, size = 1, alpha = 0.8) +
  geom_point(aes(x = year), size = 3, alpha = 0.8) +
  scale_color_viridis_d() +
  labs(x = "Year",
       y = "Count",
       color = "Category") +
  theme_minimal_custom +
  theme(panel.background = element_rect(fill = "#f2eada"),  #set background color
        axis.text = element_text(face = "bold"),              #bold axis text
        axis.title = element_text(face = "bold"),
        panel.grid.major = element_line(color = "gray"),      #set color of major grid lines
        panel.grid.minor = element_line(color = "lightgray")) #set color of minor grid lines
```

The data provided in [@fig-tab-one] and [@fig-one] offers a comprehensive insight into the dynamic changes in the homeless population within the Toronto shelter system from 2018 to 2023. It emphasizes the various transitions experienced by individuals reliant on shelter services and provides insights into the overall fluctuations in homelessness numbers.
A notable trend observed is the fluctuation in the annual count of newly identified individuals entering the shelter system. Peaking in 2018 and gradually declining over the following years, the trend suggests a persistent need for shelter services, with a gradual increase observed from 2020 onwards.
Another significant aspect of the data is the return of individuals from permanent housing to the shelter system. This phenomenon is concerning as it indicates potential housing instability for a portion of the population. Fortunately, the numbers have remained relatively stable from 2018 to 2023, and they are lower than those transitioning to permanent housing each year. However, the declining trend in individuals moving to stable housing from 2018 to 2021 is worrisome. Although there is a slight increase in 2022 and 2023, it remains below the 2018 levels, suggesting a decrease in individuals transitioning to permanent housing.
The data also captures changes in individuals returning to shelters and those becoming inactive within the shelter system. While there is a gradual decline in the number of individuals becoming inactive homeless over the five years, the decrease is minimal, similar to the trend in individuals returning to shelters. Without further explanation, it's challenging to draw reliable conclusions on the decrease in the number of inactive homeless individuals or the reasons behind individuals returning to shelters since their experience after leaving the shelter system is unknown. They could be moving to other cities, finding housing, or dying. Therefore, if there's no further detail for the reason they left the shelter system, it is hard to understand the implications of the data fully, and no conclusion can be drawn.
Lastly, individuals from the actively homeless category highlight the ongoing demand for shelter services and support. Although there was a decline in the number of actively homeless individuals in 2020 and 2021, it surged to new highs in 2022 and 2023. This suggests a significant increase in the number of individuals utilizing overnight shelter services in the past two years, indirectly indicating a potential increase in Toronto's homeless population from 2022 to 2023.
In conclusion, these data paint a complex picture of homelessness within the Toronto shelter system, with a noticeable trend of increasing actively homeless individuals. This underscores the importance of continued efforts to provide comprehensive support and housing solutions to meet the diverse needs of this vulnerable population.

## Composition of Homeless Population Groups and Correcponding Age Distribution
```{r}
#| label: fig-two
#| fig-cap: Composition of Homeless Population Groups and Age Distribution within the Toronto Shelter System between 2018 and 2023
#| echo: false
#| message: false

#### generate the 2nd graph ####
#filter for necessary population groups
cleaned_data_figtwo <- cleaned_data_figtwo %>%
  filter(population_group %in% c("All Population", "Chronic", "Refugees", "Families", "Indigenous", "Non-refugees"))

#reshape the data for graph generating
reshaped_data <- cleaned_data_figtwo %>%
  pivot_longer(cols = ageunder16:age65over, names_to = "age_group", values_to = "count")

#define a custom color palette for better look of the graph
starry_night_palette <- c("#145b7d", "#70a19f", "#008792","#00ae9d", "#76becc", "#94d6da")

#plot the data
ggplot(reshaped_data, aes(x = age_group, y = count, fill = population_group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.8) +
  labs(x = "Age Group", y = "Population Count", fill = "Population Group") +
  scale_fill_manual(values = starry_night_palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Before beginning the analysis of [@fig-two], there are several points that need to be explained. Firstly, within the population groups depicted in this graph, apart from refugees and non-refugees, the other options are not mutually exclusive. This means that a person can simultaneously belong to the chronic, families, and refugee groups, but cannot be a refugee and a non-refugee at the same time. This explains why the total number of homeless individuals in these population groups exceeds the number represented by the "all population" category. Secondly, as mentioned earlier, the number of individuals in the indigenous group was only recorded starting from 2020, so the actual number of indigenous individuals should be higher than what is shown in the graph.

We can see that individuals aged between 25 and 44 constitute the largest portion of the homeless population, while those aged between 45 and 64 represent the second largest group, and those aged 65 and above represent the smallest group. From the perspective of population groups, non-refugees constitute the largest group, followed by refugees and then chronic individuals, with indigenous individuals being the smallest group. From this distribution trend, we can infer the following information:

Firstly, except for the age group under 16, chronic individuals constitute a significant portion of the homeless population in other age groups, indicating that a considerable number of homeless individuals rely on overnight shelter services for an extended period and find it difficult to secure permanent housing, thus perpetuating their homelessness.

Secondly, although refugees constitute a smaller proportion of the homeless population compared to non-refugees overall, in the age group under 16, the number of refugees far exceeds the proportion of non-refugees. This suggests that there are a significant number of parenting teenagers or even younger children among the refugee population who are homeless. Additionally, we observe that the proportion of families in the age group under 16 also significantly increases, slightly surpassing the proportion of refugees and almost equaling the "all population" category. This evidence confirms our previous conclusions and supports the notion that these young refugee homeless individuals are likely to be wandering alongside their refugee family members.

Thirdly, unlike refugees, homeless individuals categorized as non-refugees are mainly distributed in the age group of 25 and above. Particularly in the age group of 65 and above, the number of non-refugees is almost equivalent to the "all population" category.

## Gender Composition Across Homeless Population Groups

```{r}
#| label: fig-thr
#| fig-cap: Trends in Gender Composition of Homeless Population Across Different Population Groups within the Toronto Shelter System between 2018 and 2023
#| echo: false
#| message: false

#### generate the 3rd graph ####

# Rename the columns first
cleaned_data_figthree_renamed <- cleaned_data_figthree %>%
  rename(
    "Year" = year,
    "Population Group" = population_group,
    "Male" = gender_male,
    "Female" = gender_female,
    "Transgender/Non Binary/Two spirit" = gender_transgender_non_binary_or_two_spirit
  )

# Reshape the data for graphing
cleaned_data_figthree_long <- tidyr::pivot_longer(cleaned_data_figthree_renamed, 
                                                  cols = c(Male, Female, `Transgender/Non Binary/Two spirit`),
                                                  names_to = "gender",
                                                  values_to = "count")

# Plot the data
ggplot(cleaned_data_figthree_long, aes(x = Year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_brewer(palette = "Paired") + 
  labs(x = "Date",
       y = "Count",
       fill = "Gender") +
  scale_y_continuous(labels = scales::comma) +  #format y-axis labels with commas
  theme_minimal() +
  facet_wrap(~ `Population Group`, ncol = 3, scales = "free_y") +  #facet by population groups with 3 columns
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  #rotate x-axis labels
        strip.text = element_text(size = 8),  #adjust the size of facet labels
        legend.position = "bottom",  #position legend at the bottom
        legend.box.spacing = unit(0.3, "lines"),  #set spacing around legend
        legend.spacing = unit(0.3, "lines"))  #set spacing within legend
```

The data presented in [@fig-thr] provides insights into the gender dynamics within Toronto's homeless population across different population groups and years spanning from 2018 to 2023. A consistent trend observed throughout the years is the higher representation of males compared to females and individuals identifying as transgender, non-binary, or two-spirit. This imbalance suggests a prevalent gender disparity among those experiencing homelessness, with males comprising a larger proportion of the homeless population.

However, within specific population groups, variations in gender composition emerge. For instance, in the "Chronic" population group, which likely encompasses individuals experiencing long-term homelessness, both males and females are prominently represented, indicating a diverse demographic within this subgroup. In contrast, the "Families" population group, which likely consists of homeless families with children, demonstrates a higher count of females compared to males and individuals identifying as transgender, non-binary, or two-spirit. This gender distribution within homeless families may be influenced by factors such as caregiving responsibilities and access to support services, highlighting the unique dynamics within this subgroup. Similarly, the "Refugees" population group exhibits a relatively balanced distribution between males and females, with a smaller count of individuals identifying as transgender, non-binary, or two-spirit. This suggests a diverse composition within refugee populations experiencing homelessness, possibly influenced by factors such as migration patterns and resettlement challenges.

Furthermore, the data reveal temporal fluctuations in gender composition, particularly evident in the chronic population group, which demonstrates a consistent increase in count from 2018 to 2023. This upward trend suggests a growing number of individuals experiencing long-term homelessness, underscoring the persistent challenges faced by Toronto's homeless community.

# Discussion


## Weaknesses

## Further Study



\newpage
# bibliography








